STAT 672 Final Project: Stochastic Gradient Descent

### 2018-04-15 paper/paper.tex
progress

\subsection{Basic setup}

Loosely following the set-up of \cite{bottou2010large}, consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
We have a hypothesis class $\mathcal{F}$ consisting of
various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
expresses the cost of mis-classification. Assume for now that $L$ is convex. 
We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

Because $L$ is convex, basic calculus tells us that there is one critical point,
it is the global minimum, and it is located where the gradient of the loss
function with respect to $\bm{w}$, $\nabla_{\bm{w}}$, is zero. GD is an iterative
algorithm to numerically approximate this point:

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}

$t$ refers to iterations. $\gamma$ is a parameter controlling the step
size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. 
The algorithm stops when the gradient is
arbitarily close to zero; or, put differently, when the difference between this
iteration's estimate and the last iteration's estimate is arbitrarily small.

\begin{equation}
	\bm{w}_t - \bm{w}_{t-1} \leq \epsilon
\end{equation}

Note that the batch GD algorithm presented in (2) requires evaluating the
gradient for every single observation $i$. In a large-$n$ dataset, this can be
computationally infeasible. SGD's innovation is to instead evaluate only a
single randomly chosen observation $i$ at each iteration $t$, greatly lightening
the computational load.

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma 
	\nabla_{\bm{w}}L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}

Having outlined the basic form of SGD, we now examine some of its key
properties.

@@

@@

@@

The basic SGD algorithm has been extended in different ways. The popularity of
the algorithm disallows a comprehensive or detailed treatment of all
development. This sub-section covers some of the more interesting extensions.

\cite{bottou2012stochastic}
\cite{boyd2004convex}
\cite{dal2015calibrating}
\subsubsection{Step Size (Learning Rate)}

\cite{shalev2011pegasos}

\subsubsection{Momentum}

\cite{polyak1992acceleration}
\cite{nesterov}

@@

\subsection{SGD and Statistical Learning}

\section{Conclusion}

>>>>

\subsection{Basic Form}

This section presents the basic form of GD and SGD to set up a more detailed
examination in subsequent sections. Suppose we want to minimize a function $f: \mathbb{R}^D \to \mathbb{R}$.

\begin{equation}
	\min_w f(w)
\end{equation}

$f$ takes as input $w \in \mathbb{R}^D$. Assume
that $f$ is convex and differentiable, i.e. we can compute its gradient with
respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (1),
is denoted $\hat{w}$. The iterative GD algorithm for finding $\hat{w}$ is:

\begin{equation}
	w^{(t+1)} = w^{(t)} - \gamma \nabla f(w^{(t)})
\end{equation}

$t$ refers to a particular iteration. Assume that we have supplied an initial
starting guess for $w$ for $t=0$, $w^{(0)}$. $\gamma$ refers to step size (also called
learning rate). Assume for now that $\gamma$ is fixed. The GD algorithm iterates
until some stop condition is met. This may be a fixed number of iterations, or
that the quality of approximation meets some predefined threshold. 
A common stopping condition is when the L2 norm of the gradient
is less than some arbitrarily small constant.

\begin{equation}
	||\nabla f(w^{(t)})||_2 \leq \epsilon
\end{equation}

Consider a modified situation. Suppose $f$ now takes two arguments, $w$ and $X
\in \mathbb{R}^D$. $f : \mathbb{R}^D \times \mathbb{R}^D \rightarrow
\mathbb{R}$. We have $n$ observations of $X$, and denote by
$\bm{X}_{n \times D}$ the matrix of these observations, with $X_i$ being a
particular observation or row in that matrix. We apply
$f$ over all observations, i.e. $\frac{1}{n} \sum_{i=1}^n f(w, X_i)$. Our new
problem is:

\begin{equation}
	\min_w  \frac{1}{n} \sum_{i=1}^n f(w, X_i)
\end{equation}

We could apply the GD algorithm above to find the optimal $w$.

\begin{equation}
	w^{(t+1)} = w^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n \nabla f(w^{(t)},
	X_i)
\end{equation}

But, note that doing so requires evaluating the gradient at every single
observation $i \leq n$. This may be computationally intractable for large $n$
and high-dimensional $D$. The innovation of SGD is to instead evaluate only a
single randomly-chosen $i$ at each iteration $t$.

\begin{equation}
	w^{(t+1)} = w^{(t)} - \gamma \nabla f(w^{(t)}, X_i)
\end{equation}

As it turns out, SGD converges to $\hat{w}$ and does so in a computationally
advantageous way compared to GD.

@@

This section gives both intuition and (partial) proof for why (S)GD converges to
$\hat{w}$.

Intuitively, our assumption that $f$ is convex means that there is one critical
point and it is the global minimum. Basic calculus tells us that this critical
point is located where $f'=0$ (in one dimension) or $\nabla f = 0$ (in higher
dimensions). So our task is to search in the space of $f$ for that point. We
start with some initial guess $w^{(0)}$, and every iteration, move ``down'' the
gradient in search of zero. Every iteration, we check if the gradient is equal
to 0; if not, we keep moving ``down.'' If the gradient is arbitrarily close to
zero, then we have found the critical point, which must be the value of $w$ that
minimizes $f$ and hence is $\hat{w}$ (or at least a close approximation to it).

For a more formal proof---following that of \cite{tibs_notes}---assume (in addition to previously stated assumptions
that $f$ is convex and differentiable) that the gradient of $f$ is Lipschitz-continuous
with constant $L$, i.e. that $||\nabla f(x) - \nabla f(y)||_2 \leq L||x-y||_2$
for any $x, y$. This implies that $\nabla^2 f(w) - LI$ is a negative
semi-definite matrix. We perform a quadratic expansion of $f$ around $f(x)$ and
obtain the following inequality:
$$
f(y) \leq f(x) + \nabla f(x)^T(y-x) + \frac{1}{2}\nabla^2 f(x)||y-x||_2^2
$$
\begin{equation}
	\leq f(x) + \nabla f(x)^T(y-x) + \frac{1}{2}L||y-x||_2^2
\end{equation}

Now, suppose we use the GD algorithm presented in (2) with $\gamma \leq
\frac{1}{L}$. Denote $x^+ = x - \gamma \nabla
f(x)$ and substitute $x^+$ in for $y$:

$$ f(x^+) \leq f(x) + \nabla f(x)^T(x^+ - x) + \frac{1}{2}L||x^+ - x||_2^2 $$
$$ = f(x) + \nabla f(x)^T(x - \gamma \nabla f(x) - x) + \frac{1}{2}L||x - \gamma \nabla f(x) - x||_2^2 $$
$$ = f(x) - \nabla f(x)^T \gamma \nabla f(x)  + \frac{1}{2}L||\gamma \nabla f(x)||_2^2 $$
$$ = f(x) - \gamma||\nabla f(x)||_2^2 + \frac{1}{2} L \gamma^2||\nabla f(x)||_2^2$$
\begin{equation}
= f(x) - (1 - \frac{1}{2}L\gamma)\gamma||\nabla f(x)||_2^2 
\end{equation}

We have defined $\gamma \leq \frac{1}{L}$. This implies:
$$
-(1 - \frac{1}{2}L \gamma) = \frac{1}{2}L \gamma - 1
\leq \frac{1}{2}L\frac{1}{L} - 1
= \frac{1}{2} - 1
= - \frac{1}{2}
$$

Returning to (8), we obtain:

\begin{equation}
f(x^+) \leq f(x) - \frac{1}{2}\gamma||\nabla f(x)||_2^2
\end{equation}

By definition, a squared L2 norm will always be positive unless its content is
0, and we have defined $\gamma$ to be positive, so $\frac{1}{2} \gamma||\nabla f(x)||_2^2$ will always be positive unless the
gradient is equal to zero. So, (9) implies that GD results in a strictly
decreasing objective function value until it reaches the point where the
gradient equals zero, which is the optimal value. A key caveat is an
appropriately chosen $\gamma$, a point covered in more detail later.

The above proof is for GD. A rigorous proof is not given for why SGD also
converges to the optimal value. Informally, we can note that the above proof
says given infinite $t$ and appropriate $\gamma$, the iterative algorithm will
always converge to the optimal value. SGD is doing the same thing as GD, 
just with a single observation per iteration 
rather than the entire dataset per iteration. Since SGD is using less information
per iteration, we would expect the convergence to require more iterations. But
per (9) it is guaranteed to \textit{eventually} arrive at the optimal solution.
Put differently, the difference between GD and SGD must then primarily revolve
around convergence speed, not the final value that is converged to.

@@

Table 1---a simplified version of that in \cite{bottou2010large}---illustrates
the computational advantages of SGD over GD. As shown in row 1, because SGD uses
less information per iteration than GD, it requires more iterations to achieve
some fixed degree of accuracy $\rho$.
However, row 2 shows that because GD computes the gradient for every observation every iteration,
while SGD only does so for a single randomly chosen iteration, GD's time per
iteration scales linearly with $n$ while SGD's is a constant. Thus, we conclude
that SGD's time to
reach accuracy $\rho$ scales only with $\rho$, while GD's scales with both
$\rho$ and---crucially---linearly with $n$. So, the larger $n$ grows, the larger
SGD's computational advantage over GD.

\begin{table}[h!]
	\centering
	\caption{Asymptotic comparison of GD and SGD}
	\begin{tabular}{|l l l|}
		\hline
		& \textbf{GD} & \textbf{SGD} \\
		Iterations to accuracy $\rho$ & $\log \frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
		Time per iteration & $n$ & 1 \\
		Time to accuracy $\rho$ & $n \log \frac{1}{p}$ & $\frac{1}{\rho}$ \\
		\hline
	\end{tabular}
\end{table}

We can visually depict SGD's ``noisier'' convergence in comparison to GD.

@@

The basic SGD algorithm has been extended in many different ways. The popularity of
the algorithm disallows a comprehensive or detailed treatment of all
development. This sub-section covers some of the more important and interesting extensions.

\subsubsection{Step Size}

As hinted at in our proof of convergence to $\hat{w}$, much depends on
an appropriately chosen step size $\gamma$. If $\gamma$ is too small, (S)GD may
take an excessively long time to run (since we are only moving a small distance
``down'' the gradient every iteration); if $\gamma$ is too large, then an
iteration's step may ``overshoot'' the critical point and prevent convergence.
An entire literature has developed around the best way to select $\gamma$. The
central insight that is common to most $\gamma$-selection methods is to treat it
as a time-indexed rather than fixed hyperparameter, i.e., $\gamma^{(t)}$ rather
than $\gamma$. Ideally we would like $\gamma^{(t)}$ to be large when far away
from the critical point (so as to increase speed of convergence) and to be small
when close to the critical point (so as to avoid overshooting).

Line search is one method for computing step size. As described in
\cite{boyd2004convex}, there are two main variants: \textit{exact} and
\textit{backtracking}. In exact line search, $\gamma$ is chosen to minimize $f$
along the ray $\{x + \gamma \Delta x\}$, where $\gamma \in \mathbb{R}^+)$ and
$\Delta x$ is the descent direction determined by SGD:
\begin{equation}
	\gamma = \argmin_{s \geq 0} f(x + s\Delta x)
\end{equation}
This obviously is an optimization problem in and of itself, and
so it can be computationally impractical to add this burden in addition to the
``main'' optimization problem we are trying to solve using SGD. For this reason,
\textit{backtracking} is an iterative approximation method that is
computationally lighter. The backtracking algorithm takes two hyper-parameters
$\alpha \in (0, 0.5)$ and $\beta \in (0, 1)$. It starts with an initial guess of
$s^{(0)} = 1$, and then for every iteration $t$, updates $s := \beta t$. The
algorithm stops when $f(x + s \Delta x) \leq f(x) + \alpha s \Delta f(x)^T
\Delta x$, and the final value of $s$ is taken as $\gamma$.

\cite{bottou2012stochastic} advocates using learning rates of the form
$\gamma^{(t)} = \gamma^{(0)}(1 + \gamma^{(0)}\lambda t)^{-1}$. When the Hessian
$H$ of $f$ is strictly positive, using $\gamma^{(t)} =
(\lambda_{\mathrm{min}}t)^{-1}$, where $\lambda_{\mathrm{min}}$ is the smallest
eigenvalue of $H$, produces the best convergence speed. Note that $
(\lambda_{\mathrm{min}}t)^{-1}$ decreases asymptotically with $t$, matching our intuition
about larger step sizes in early iterations and smaller step sizes in later
iterations. However, simply using $\gamma^{(t)} =
(\lambda_{\mathrm{min}}t)^{-1}$ can produce \textit{too} large of steps in early
iterations. Hence, it often works better to start with some reasonable initial
estimate $\gamma^{(0)}$ that then decays in the fashion of
$(\lambda_{\mathrm{min}}t)^{-1}$, leading to the expression in the first
sentence of this paragraph. By definition, ``reasonable'' is more of a judgment
call than a mathematical statement: Bottou provides some heuristics for creating
such an estimate of $\gamma^{(0)}$.

\subsubsection{Momentum and Acceleration}

In our set-up of SGD, we stipulated that the function to be minimized, $f$, is
convex. Although this is a necessary condition for some proofs of SGD's
correctness, SGD is widely used in applications---most prominently, neural
networks and deep
learning---where this assumption is not
always true (as covered in more detail in CHAPTER X). In such settings, it is
possible that there are multiple local minima. The challenge for SGD is
how to avoid getting ``stuck'' in these local ravines and continue iterating
until the true global minimum is achieved.

Momentum---also called acceleration---is a common technique for avoiding this. As outlined in
\cite{rumelhart1986general} and 
\cite{qian1999momentum}, we add a new acceleration term to the familiar GD
equation. Let $z^{(t+1)} = \alpha z^{(t)} + \nabla f(w^{(t)})$. GD with
momentum then is:

\begin{equation}
	w^{(t+1)} = w^{(t)} - \gamma z^{(t+1)}
\end{equation}

Now, the weight vector obtained for the current timestep depends both on the
current gradient \textit{and the gradient of the previous time-step}.
Intuitively, without acceleration, SGD tends to oscillate within a ravine
because the gradient is dominated by the steep curves on either side of the
ravine; acceleration helps point the descent along the shallower curve leading
out of the ravine. \cite{nesterov} provides another popular implementation of
acceleration.
\cite{polyak1992acceleration}

@@

\cite{shalev2011pegasos}
\subsection{SGD and Statistical Learning}
\cite{dal2015calibrating}

Loosely following the set-up of , consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D, n \in \mathbb{R}^n, D \in \mathbb{R}^D}$ (corresponding to $n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict a particular observation's label $Y_i$ using that observation's features $\bm{X}_i$. 
We have a hypothesis class $\mathcal{F}$ consisting of
various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
expresses the cost of mis-classification. Assume for now that $L$ is convex. 
We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

Because $L$ is convex, there is one critical point,
it is the global minimum, and it is located where the gradient of the loss
function with respect to $\bm{w}$ is zero. GD is an iterative
algorithm that can be used to numerically approximate this point. We first state
it in a general form, and then adapt it to the specific problem outlined above.

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}

$t$ refers to iterations. $\gamma$ is a parameter controlling the step
size (also called ``learning rate''). Assume for now that $\gamma$ is fixed. For
$t=0$ we must supply an initial starting guess, $\bm{w}_0$.

\begin{equation}
	\left|\frac{1}{n}\sum_{i=1}^n \nabla L(\bm{Y}_i,
	f_{\bm{w}_t}(\bm{X}_i))\right| \leq \epsilon
\end{equation}

Note that the batch GD algorithm presented in (2) requires evaluating the
gradient for every single observation $i$. In a large-$n$ dataset, this can be
computationally infeasible. SGD's innovation is to instead evaluate only a
single randomly chosen observation $i$ at each iteration $t$.

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma 
	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}
\section{Conclusion}
